\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Understanding Driver Behavior: Studies on NYC Taxi Data}

### 2015-05-11 Report/article_2.tex
edited report

\author{
\large
\textsc{Daniel Abramson}\\[2mm] % Your name
\normalsize New York University \\ % Your institution
\normalsize \href{mailto:dna237@nyu.edu}{dna237@nyu.edu} % Your email address
\vspace{-5mm}
\and
\textsc{Yoon Kim}\\[2mm] % Your name
\normalsize New York University \\ % Your institution
\normalsize \href{mailto:yhk255@nyu.edu}{yhk255@nyu.edu} % Your email address
\vspace{-5mm}
\and
\textsc{Sylvia Zhao}\\[2mm] % Your name
\normalsize New York University \\ % Your institution
\normalsize \href{mailto:ssz8@cornell.edu}{ssz8@cornell.edu} % Your email address
\vspace{-5mm}
}
\date{}

@@

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------

\begin{abstract}

\noindent Using all data on NYC taxi trips and fares from 2010 and 2013 we perform a series of studies to understand motivating factors behind driver behavior as well as outcomes from those incentives. In order to analyze this large dataset ($\approx$14GB), we employ a number of big data processing technologies that work on the Hadoop framework: Spark and MapReduce programs.  We present our findings, issues encountered, and experimental setup. Of these findings, the most notable include: there are outlier drivers who are on average more productive per day, the sensitivity of fares to environment variables are seemingly invariant, the distribution of actual to predicted fares implies a number of overcharges that should be investigated, drivers can expect better tips from passenger drop-off to certain areas, and previous tips affect driver behavior for the next trip.

@@

We ask a series of questions that are pertinent to both lenses. 
\begin{enumerate}
\item Are there productivity differences among drivers? Do some drivers consistently make more revenue per unit time worked than others? 
\item Can we predict the sensitivities of taxi fares to a number of different environmental variables that would affect driver incentives to work? These include hours of the day, days of the week, and seasonal factors.
\item Using this model, can we identify ``outlier" drivers whose actual fare exceeds the predicted fare (i.e. those that are overcharging)? How is tip behavior affected by such things?
\item Do previous earnings affect driver behavior? Is there some sort of 'hot-hand' with respect to tips? If a driver receives an above-median tip does this affect his behavior on the next ride? Is he motivated to continue good performance or does he care less? Is there no change in behavior?
\item How is tip affected by the income of the passengers, as measured by the median household income of the zip code in which the passenger is dropped off?

\end{enumerate}

@@

We used AWS to perform this join, first transferring all input files to S3. We then used 4 m3.xlarge machines to perform the join. This took about 2 hours for each. There were 7 reducers. Mapper numbers were set to defaults.

\section{Investigating Question 1: Driver Productivity Differences}

@@

\subsection{Results}
\includegraphics[scale=.35]{Minutes_v_Rev.png}

Though there is nothing surprising about a positive linear relationship between Daily Revenue and Minutes Worked per day, we find the outliers to be interesting. There seems to be a larger spread of outliers below the line than above. However, analyzing the outliers above the line, some seem to work very few minutes an contribute a lot to revenue. Perhaps these drivers are not properly incentivized. It would appear they have much downtime during the day.

@@

\section{Investigating Question 3: Driver Performance v Fare Sensitivities}
Based on the regression output from the previous section, we plot the residual (i.e. actual fare minus predicted fare) for each trip and bin them into histograms.  
\includegraphics[scale=.35]{figure_1.png}
While much the error may be coming from the fact that our regression model is very simple (for example, it doesn't take into account the fact that a distance of 1 mile means very different things in midtown versus in suburbs of Queens), there does seem to be a cohort of trips in which the customer is overcharged. \\
Next we investigate whether customers who are overcharged (i.e. have high residuals) give less tip. Below is a scatter plot of residuals (x-axis) versus tip percentage (y-axis). While the relationship is far from clear, there does seem to be a slight negative relationship between tip percentage and residual.
\includegraphics[scale=.35]{figure_2.png}

@@

\includegraphics[scale=.4]{zip_income.png}
As can be seen from the above chart, there seems to be a very small positive relationship between median household income and average tip percentage.
%------------------------------------------------

\section{Concluding Remarks}

@@

>>>>

\author{
\large
\textsc{Daniel Abramson}\\[2mm] % Your name
\normalsize New York University \\ % Your institution
\normalsize \href{mailto:dna237@nyu.edu}{dna237@nyu.edu} % Your email address
\vspace{-5mm}
\and
\textsc{Yoon Kim}\\[2mm] % Your name
\normalsize New York University \\ % Your institution
\normalsize \href{mailto:yhk255@nyu.edu}{yhk255@nyu.edu} % Your email address
\vspace{-5mm}
\and
\textsc{Sylvia Zhao}\\[2mm] % Your name
\normalsize New York University \\ % Your institution
\normalsize \href{mailto:ssz225@nyu.edu}{ssz225@nyu.edu} % Your email address
\vspace{-5mm}
}
\date{}

@@

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------
\begin{abstract}
\noindent Using all data on NYC taxi trips and fares from 2010 and 2013 we perform a series of studies to understand motivating factors behind driver behavior as well as outcomes from those incentives. In order to analyze this large dataset ($\approx$14GB), we employ a number of big data processing technologies that work on the Hadoop framework: Spark and MapReduce programs.  We present our findings, issues encountered, and experimental setup. Of these findings, the most notable include: tip aggregates and operating taxis are sensitive to day of the week, there are outlier drivers who are on average more productive per day, the sensitivity of fares to environment variables are seemingly invariant, the distribution of actual to predicted fares implies a number of overcharges that should be investigated, drivers can expect better tips from passenger drop-off to certain areas, and previous tips affect driver behavior for the next trip.

@@

We ask a series of questions that are pertinent to both lenses. 
\begin{enumerate}
\item Are there productivity differences among drivers? Do some drivers consistently make more revenue per unit time worked than others? 
\item Can we predict the sensitivities of taxi fares to a number of different environmental variables that would affect driver incentives to work? These include hours of the day, days of the week, and seasonal factors.
\item Using this model, can we identify ``outlier" drivers whose actual fare exceeds the predicted fare (i.e. those that are overcharging)? How is tip behavior affected by such things?
\item Do previous earnings affect driver behavior? Is there some sort of 'hot-hand' with respect to tips? If a driver receives an above-median tip does this affect his behavior on the next ride? Is he motivated to continue good performance or does he care less? Is there no change in behavior?
\item How is tip affected by the income of the passengers, as measured by the median household income of the zip code in which the passenger is dropped off?
\item Do passengers tip more with respect to methods of payment and number of riders? Is driver behavior affected by tipping trends?
\end{enumerate}

@@

We used AWS to perform this join, first transferring all input files to S3. We then used 4 m3.xlarge machines to perform the join. This took about 2 hours for each. There were 7 reducers. Mapper numbers were set to defaults. 
%------------------------------------------------
\section{Investigating Question 1: Driver Productivity Differences}

@@

\subsection{Results}

Though there is nothing surprising about a positive linear relationship between Daily Revenue and Minutes Worked per day, we find the outliers to be interesting. There seems to be a larger spread of outliers below the line than above. However, analyzing the outliers above the line, some seem to work very few minutes an contribute a lot to revenue. Perhaps these drivers are not properly incentivized. It would appear they have much downtime during the day. 
\includegraphics[scale=.35]{Minutes_v_Rev.png}

@@

\section{Investigating Question 3: Driver Performance v Fare Sensitivities}
Based on the regression output from the previous section, we plot the residual (i.e. actual fare minus predicted fare) for each trip and bin them into histograms.  
\includegraphics[scale=.34]{figure_1.png}
While much the error may be coming from the fact that our regression model is very simple (for example, it doesn't take into account the fact that a distance of 1 mile means very different things in midtown versus in suburbs of Queens), there does seem to be a cohort of trips in which the customer is overcharged. \\
Next we investigate whether customers who are overcharged (i.e. have high residuals) give less tip. Below is a scatter plot of residuals (x-axis) versus tip percentage (y-axis). While the relationship is far from clear, there does seem to be a slight negative relationship between tip percentage and residual.
\includegraphics[scale=.345]{figure_2.png}

@@

\includegraphics[scale=.35]{zip_income.png}
As can be seen from the above chart, there seems to be a very small positive relationship between median household income and average tip percentage.
%------------------------------------------------

\section{Investigating Question 6: Aggregate Trends in Tipping and Driver Behavior}
\subsection{Experimental Setup}
A series of MapReduce functions were utilized to generate aggregate trend visualizations of fares and the combined trips and fares data from 2010. Tipping and driver behaviors were examined using the following variables: number of passengers per trip, day of the week, and method of fare payment.

\subsection{Results}
To obtain an overview of tipping trends, the ratio of the count of rides with positive tips over all rides was plotted against the methods of payment. Percentage tip was calculated over all data as well as over data with tip equalling zero dropped.
\includegraphics[scale=.35]{pct_pc0_pcno0.pdf}

The graph points to a nearly absolute drop in the frequency of tipped trips when the payment method is not under automated reporting systems (i.e. credit cards). Filtering out zeros in tip records, the plot shows a fairly consistent rate of 20 percent in tips which agrees with the more reliably reported complete credit cards tip data. Therefore, it is unlikely that the drastic drop in frequency of tipped trips is due to passenger behavior changed based on payment type. This highlights the problem of drivers "cheating" on reporting approximately 96 percent of instances of non-credit card tips, making non-credit card tip data unreliable in raw form. This problem may be an issue for taxi companies when calculating compensation packages for contracted drivers.

Since non-credit card tip data was determined to be unreliable, the plot of number of passengers against tip percentage was generated over only trips paid with credit cards.
\includegraphics[scale=.35]{tip_passenger.pdf}
Other than outliers of 0, 7, and 9 passengers, the tip percentage is fairly consistent at just slightly below 20 percent, which is in agreement with the aggregate statistic in the previous graph. Therefore, number of passengers is not supported as a source of deviation from normal tipping behavior.
Overall, passenger tipping behavior is determined to be inelastic to payment type and number of taxi riders.

To examine the effects of day of the week on tip and driver behavior, the total sum of tips and unique taxis were plotted against days.
\includegraphics[scale=.32]{total_tip.pdf}
\includegraphics[scale=.32]{total_taxis.pdf}
Both graphs peak on Thursday and drop sharply on the weekends. With the exception of Monday and Tuesday, the graphs demonstrate similar trend shape. While other variables may be in play at the micro scale, the aggregate view supports a positive correlation between higher tips and willingness of drivers to participate in work.

%------------------------------------------------
\section{Concluding Remarks}

@@

\subsubsection*{Sylvia}
\begin{itemize}
  \item Question 6: Aggregate Trends in Tipping and Driver Behavior
  \item Presentation - Writing
  \item Report - Writing
  \item Code - Organization for Submission
\end{itemize}
